SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2013-02-22 paper/paper.tex
updates for paper.

% can be 9pt
\documentclass[10pt]{sigplanconf}
\usepackage{url}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{color}
\usepackage{url}
\usepackage{alltt}
\usepackage{tabularx}
\usepackage{subfigure}
\usepackage{fancyhdr}
\usepackage{paralist}

@@

\newenvironment{code}{\begin{alltt}\small}{\end{alltt}}
%% \newenvironment{cols}{\begin{tabular}{m{3.6cm}|m{3.6cm}}{Haskell} &

@@

In industrial uses, shrinking is essential.  Hughes, an inventor of QuickCheck
and the CEO of Quiviq, a company that commercializes the technology, has noted
that ``without it [shrinking], randomly generated failing cases would often be
so large as to be almost useless.''~\cite{qcjh}.  Hughes~\emph{et al.} give an
extended example in which shrinking is essential in debugging telecom
software~\cite{telecom}.

@@

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
           & QC (none)  & QC (trunc) & QC (tuple) & SmartCheck\\
\hline \hline
Mean       & 0.08s, 69  & 0.53s, 35  & 2.24s, 12  & 0.10s, 7\\
\hline
Median     & 0.07s, 69  & 0.17s, 33  & 1.81s, 13  & 0.10s, 6\\
%% \hline
%% MAD        & 0.02s, 13  & 0.09s, 8   & 0.62s, 3   & 0.02s, 2\\
\hline
95\%       & 0.14s, 100 & 1.76s, 65  & 4.06s, 19  & 0.18s, 12\\
\hline
    \end{tabular}
  \end{center}
  \caption{Results for 1000 tests of the program in Figure~\ref{fig:initial}.}
  \label{table:results}
\end{table}

\todo{get 99\% }

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider
Table~\ref{table:results}, giving the results for using QuickCheck and
SmartCheck on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck,
since as noted, it is not even feasible for the example).  Each entry is a pair
containing the number of seconds required for counterexample discovery and
shrinking and the size of the final value obtained (by counting \ttp{N}
constructors).  We show the mean, median, median, and the values at the 95th
percentile values.  (The curves produced by these values are not normal, so we
omit standard the deviation.)

In the first three columns, the results of using QuickCheck to generate
counterexamples are shown.  The first column contains the results of using
QuickCheck without shrinking to provide a baseline.  The next column shows the
result of using QuickCheck with truncating shrink, which each list is truncated
to 10 elements.  Compared with no shrinking, this approximately halves the size
of the median counterexample, but doubles the time taken.  There can be large
outliers, however, in the time taken during shrinking.  Next, we show the
results of the ``tuple shrink'' instance.  While the median amount of time
required is much higher, the size of the resulting value is approximately halved
again.  The reason for the increase in time is because shrinking is very
productive and produces a nonempty list for most values.

In the last column are the results for SmartCheck, which includes the time
required by QuickCheck to generate the original counterexample.  The median time
required by SmartCheck to shrink is 0.02 seconds (SmartCheck's execution time
minus the execution time for QuickCheck without shrinking), and the median size
of a resulting value is six \ttp{N} constructors.  Moreover, SmartCheck has a
very small variance---note the mean and median are nearly identical, meaning
there are very few outliers.  And we stress that SmartCheck achieves these
results \emph{generically}, requiring no shrink definitions to be provided by
the user.

@@

For example, consider a simple calculator language containing constants,
addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\noindent
Consider a program that takes an \ttp{Exp} as an argument that throws an error
if an evaluation of an \ttp{Exp} value contains a division-by-zero.  Testing for
the occurrence of the error, we might have a counterexample like the following:
%
\medskip%
\begin{sloppypar}
\noindent%
\ttp{Add (Div (C 5) (C (-12))) (Add (Add (C 2) (C 4)) (Add (C 7) (Div (Add (C 7)
  (C 3)) (Add (C (-5)) (C 5)))))}
\end{sloppypar}
\medskip%
In the counterexample, the culprit is a buried sub-value,
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
Thus, when attempting to shrink an \ttp{Exp} value, it pays to test whether a
sub-value itself fails the property.

@@

\todo{xmonad}
\todo{use SC to test itself?}
\todo{experiments? scale vs. broad data, linear data?}

@@

HDD is most similar to SmartCheck's reduction algorithm, with an important
difference: HDD (and DD) is deterministic, so the algorithm only succeeds in
reducing the counterexample only if a new counterexample can be constructed from
the original one.  Our approach combines the speed of delta debugging with the
power of QuickCheck to \emph{randomly} discover structurally smaller
counterexamples.  Furthermore, neither paper explores the idea of counterexample
generalization, however.

@@

We have benchmarked Feat as well on the example in the introduction
(Figure~\ref{fig:initial}).  While it performs quite competitively to SmartCheck
in terms the size of counterexamples discovered: the mean and median number of
\ttp{N} constructors in the counterexamples returned by Feat are both seven.
However, Feat performs more poorly with respect to how long counterexample
discovery takes: the mean time taken by Feat is 5.60 seconds, with a long tail:
five percent of runs take over 14 seconds (recall that the mean time taken by
SmartCheck is 0.12 seconds, and 99\% of runs take less than 0.30 seconds).

@@

\todo{shrink paper rehger pointed to ``Automatic isolation of compiler errors''
  from 1994}
\todo{\url{http://www.cs.utah.edu/~regehr/papers/pldi12-preprint.pdf}}

%% \todo{avoid complexities \url{http://stackoverflow.com/questions/14006005/idiomatic-way-to-shrink-a-record-in-quickcheck}%
%% %
%% * Also check out \url{http://www.whyprogramsfail.com/}%
%% %
%% \url{http://www.haskell.org/haskellwiki/Research_papers/Testing_and_correctness}%}

%% \todo{Not relevant really: GenCheck
%%   \url{https://github.com/JacquesCarette/GenCheck}%
\url{http://blog.regehr.org/archives/697} --- check out C-reduce (Regehr)
%% }

\todo{stuff from SC repo}

@@

>>>>

% can be 9pt
\documentclass{sigplanconf}
\usepackage{url}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{color}
\usepackage{url}
\usepackage{alltt}
\usepackage{tabularx}
\usepackage{subfigure}
\usepackage{fancyhdr}
\usepackage{paralist}

@@

\newenvironment{code}{\begin{alltt}}{\end{alltt}}
%% \newenvironment{cols}{\begin{tabular}{m{3.6cm}|m{3.6cm}}{Haskell} &

@@

In industrial uses, shrinking is essential.  Hughes, in describing industrial
uses of QuickCheck, has noted that ``without it [shrinking], randomly generated
failing cases would often be so large as to be almost useless.''~\cite{qcjh}.
Hughes~\emph{et al.} give an extended example in which shrinking is essential in
debugging telecom software~\cite{telecom}.

@@

\paragraph{SmartCheck}

\begin{figure}[ht]
\includegraphics[scale=0.6]{Graphs/data}
\includegraphics[scale=0.6]{Graphs/time-big}
\includegraphics[scale=0.6]{Graphs/time-small}
  \caption{Results for 2000 tests of the program in Figure~\ref{fig:initial}.}
  \label{fig:graphs}
\end{figure}

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
           & QC (none)  & QC (trunc) & QC (tuple) & SmartCheck\\
\hline \hline
Mean       & 0.08s, 69  & 0.53s, 35  & 2.24s, 12  & 0.10s, 7\\
\hline
Median     & 0.07s, 69  & 0.17s, 33  & 1.81s, 13  & 0.10s, 6\\
%% \hline
%% MAD        & 0.02s, 13  & 0.09s, 8   & 0.62s, 3   & 0.02s, 2\\
\hline
95\%       & 0.14s, 100 & 1.76s, 65  & 4.06s, 19  & 0.18s, 12\\
\hline
    \end{tabular}
  \end{center}
  \caption{Summarizing data for the graphs in Figure~\ref{fig:graphs}. Entries
    contain execution time (in seconds) and counterexample sizes (counting
    \ttp{N} constructors).}
  \label{table:results}
\end{table}

Motivated by these limitations of QuickCheck and SmallCheck, we developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  SmartCheck takes a counterexample produced by some
oracle---in these experiments, we use QuickCheck---and then shrinks the
counterexample.

To motivate the benefit of SmartCheck, consider the graphs in
Figure~\ref{fig:graphs} and the Table~\ref{table:results}, giving the results
for using QuickCheck and SmartCheck on the program shown in
Figure~\ref{fig:initial} (we omit SmallCheck, since as noted, it is not even
feasible for the example).  The graphs show the number of results, out of 2000
test runs, with the specified size of the final counterexample (by counting
\ttp{N} constructors) or requiring the specified execution time.  We show two
execution time graphs, since the scale of time scales differ dramatically.  In
the table, we show the mean, median, median, and the results at the 95th
percentile values.  (The curves produced by these values are not necessarily
Gaussian, so we omit standard the deviation.)

The experiments show three approaches to shrinking with QuickCheck: no
shrinking, to provide a baseline, truncated shrinking, as described above, in
which each list is truncated to 10 elements, and tuple shrinking, also described
above.  Note that SmartCheck produces smaller counterexamples than the other
approaches, with a very small tail.  SmartCheck has a very small variance---note
the mean and median are nearly identical, meaning there are very few outliers.

More dramatic, however, are the differences
in execution time.  The execution times for SmartCheck include the initial
counterexample generation time required by QuickCheck; thus the time required by
SmartCheck to shrink a counterexample is on average just 0.2 seconds (0.10 -
0.08).  As can be seen in the graphs, SmartCheck's execution time is a linear
function (like QuickCheck, without shrinking).  Truncation and tuple shrinking
produce very long tails, with some counterexamples taking over a minute to
shrink.  SmartCheck is largely insensitive to the size of the original
counterexample provided.

In the last column are the results for SmartCheck, which includes the time
required by QuickCheck to generate the original counterexample.  The median time
required by SmartCheck to shrink is 0.02 seconds (SmartCheck's execution time
minus the execution time for QuickCheck without shrinking), and the median size
of a resulting value is six \ttp{N} constructors.

And we stress that SmartCheck achieves these results \emph{generically},
requiring no shrink definitions to be provided by the user.

@@

For example, consider a simple calculator language containing constants,
addition, and division:
%
\begin{code}
data Exp = C Int
         | Add Exp Exp
         | Div Exp Exp
\end{code}
%
\noindent
Consider a program that takes an \ttp{Exp} as an argument that throws an error
if an evaluation of an \ttp{Exp} value contains a division-by-zero.  Testing for
the occurrence of the error, we might have a counterexample like the following:
%
\medskip%
\begin{sloppypar}
\small
\noindent%
\ttp{Add (Div (C 5) (C (-12))) (Add (Add (C 2) (C 4)) (Add (C 7) (Div (Add (C 7)
  (C 3)) (Add (C (-5)) (C 5)))))}
\end{sloppypar}
\medskip%
\noindent
In the counterexample, the culprit is a buried sub-value,
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
Thus, when attempting to shrink an \ttp{Exp} value, it pays to test whether a
sub-value itself fails the property.

@@

\subsection{Xmonad}
Xmonad is a popular X11 window manager written in Haskell~\cite{xmonad}.  Xmonad
is a large software project with many contributors, so consequently, a
QuickCheck test harness is included to help ensure new commits do not introduce
bugs.  At the heart of Xmonad is a \ttp{StackSet} data type that encodes the
relationship between windows, work spaces, and which window has the focus.
Xmonad contains properties to ensure the correct manipulation of
\ttp{StackSet}s.

Xmonad passes all of its QuickCheck tests, but let us see what might happen to a
new contributor if things go awry.  Suppose a developer defines an insertion
function that inserts a new focus into the \ttp{StackSet}, except she forgets to
place the value inside.  The correct insertion function from Xmonad is
%
\begin{code}
insertUp a s = if member a s then s else insert
  where
  insert = modify (Just $ Stack a [] [])
                  (\textbackslash(Stack t l r) ->
                      Just $ Stack a l (t:r)) s
\end{code}
%
\noindent
So suppose the developer forgets to cons \ttp{t} onto \ttp{r}.  Fortunately,
there are QuickCheck tests that depend on \ttp{insertUp}.  One such test is
\ttp{prop\_shift\_reversible}, which tests if the \ttp{shift} function, which
checks a property about shifting focus back and forth between windows.

With the incorrect version of \ttp{insertUp}, QuickCheck returns a wall of text
to the user, showing the fields of a \ttp{StackSet}:

\medskip%
\begin{sloppypar}
\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
  \{getNonNegative = 5\}, layout = -2, stack = Just (Stack \{focus = 'I', up =
  ", down = "\})\}, screen = 5, screenDetail = 0\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 0\}, layout =
    -2, stack = Just (Stack \{focus = '\char`\\161', up = \", down =
    \"\})\}, screen = 1, screenDetail = -2\},Screen \{workspace = Workspace
    \{tag = NonNegative \{getNonNegative = 2\}, layout = -2, stack = Just (Stack
    \{focus = '3', up = \", down = "\})\}, screen = 2, screenDetail =
    -1\},Screen \{workspace = Workspace \{tag = NonNegative \{getNonNegative =
    3\}, layout = -2, stack = Just (Stack \{focus = 'T', up = \", down =
    \"\})\}, screen = 3, screenDetail ) = 2\},Screen \{workspace = Workspace
    \{tag = NonNegative \{getNonNegative = 4\}, layout = -2, stack = Nothing\},
    screen = 4, screenDetail = -1\},Screen \{workspace = Workspace \{tag =
    NonNegative \{getNonNegative = 6\}, layout = -2, stack = Just (Stack \{focus
    = 'm', up = \", down = \"\})\}, screen = 6, screenDetail = 2\},Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 7\}, layout =
    -2, stack = Nothing\}, screen = 0, screenDetail = 0\}], hidden = [Workspace
    \{tag = NonNegative \{getNonNegative = 1\}, layout = -2, stack = Just (Stack
    \{focus = '\char`\\DC3', up = \", down = \"\})\},Workspace \{tag = NonNegative \{getNonNegative = 8\}, layout = -2, stack = Nothing\},Workspace \{tag = NonNegative \{getNonNegative = 9\}, layout = -2, stack = Just (Stack \{focus = '\char`\\ACK', up = \", down = \"\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
This is a wall of text, which provides little useful information about what went
wrong.  SmartCheck, however, takes the value and returns the following:
%
\begin{code}
Extrapolated value:
forall x0 x1 x2:
  StackSet x0 x1 x2 (fromList [])
\end{code}
%
\noindent
The programmer immediately sees that the first three fields in the record are
irrelevant,  The last field is a dictionary (i.e., an ordered map), which we
declare to be a base type, so it is not analyzed by SmartCheck.  The upshot is
that any \ttp{StackSet} fails because insertion is ill-defined.  Such an
insight, while obvious in hind-sight, can take hours to arrive at as the
programmer manually tries to write her own shrink functions or manually
deconstruct a large record.

\todo{xmonad} \todo{experiments? scale vs. broad data, linear data?}

@@

HDD is most similar to SmartCheck's reduction algorithm, with an important
difference: HDD (and DD) is deterministic, so the algorithm only succeeds in
reducing the counterexample only if a new counterexample can be constructed from
the original one.  Our approach combines the speed of delta debugging with the
power of QuickCheck to \emph{randomly} discover structurally smaller
counterexamples; in short, this is the first treatment we know of that
integrates counterexample generation with counterexample reduction.
Furthermore, neither paper explores the idea of counterexample generalization.

@@

We have benchmarked Feat as well on the example in the introduction
(Figure~\ref{fig:initial}).  While it performs quite competitively to SmartCheck
in terms the size of counterexamples discovered: the mean and median number of
\ttp{N} constructors in the counterexamples returned by Feat are both seven.
However, Feat performs more poorly with respect to how long counterexample
discovery takes: the mean time taken by Feat is 5.60 seconds, with a long tail:
five percent of runs take over 14 seconds (the mean time taken by
SmartCheck is 0.12 seconds, and 99\% of runs take less than 0.30 seconds).

@@

%% \todo{avoid complexities \url{http://stackoverflow.com/questions/14006005/idiomatic-way-to-shrink-a-record-in-quickcheck}%
%% %
%% * Also check out \url{http://www.whyprogramsfail.com/}%
%% %
%% \url{http://www.haskell.org/haskellwiki/Research_papers/Testing_and_correctness}%}
%% \todo{shrink paper rehger pointed to ``Automatic isolation of compiler errors''
%%   from 1994}

%% \todo{Not relevant really: GenCheck
%%   \url{https://github.com/JacquesCarette/GenCheck}%
%% \url{http://blog.regehr.org/archives/697} --- check out C-reduce (Regehr)
%% \todo{\url{http://www.cs.utah.edu/~regehr/papers/pldi12-preprint.pdf}}
%% }

@@

We have not profiled the performance of our implementation of SmartCheck, so
additional performance gains might be possible.  In particular, we perform
significant computations over tree-structured data, and a zipper~\cite{zipper},
which we do not currently implement, could improve performance.
